---
title: "epicoda: Introduction and Examples"
output: 
  rmarkdown::html_vignette:
      number_sections: true
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{vignette-epicoda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 
```

```{r, include = FALSE}
devtools::load_all()
```
This code chunk installs and loads the package: 
```{r, eval = FALSE}
install.packages("devtools") # To install epicoda from GitHub, the devtools package is required.
library(devtools)
devtools::install_github(
  "activityMonitoring/epicoda",
  build_opts = c("--no-resave-data"),
  build_vignettes = TRUE,
  build_manual = TRUE
)
library(epicoda)
```
# A basic analysis
This vignette illustrates some of the things this package can be used to do in an epidemiological analysis with a compositional exposure. More information about any function can  be found by running `?function_name` (as usual).

## Getting started with compositional data 
### Loading simulated data
`simdata` is a (simulated) compositional dataset in the package. It has 5 columns which make up a hypothetical physical behaviour composition: "vigorous" (activity), "moderate" (activity), "light" (activity), "sedentary" (behaviour), "sleep". It also contains sex, age group and outcome columns. As the data is simulated, all results are hypothetical. 
```{r}
head(simdata)
```

### What is compositional data? 
Data is compositional when the *relative* (not absolute) values of the variables are relevant to the problem being studied [@Filzmoser2018]. Another way to describe this is that compositional data is data that can be expressed as proportions of a whole without losing information. 

Here, the compositional variables sum to the same total (24- the number of hours in the day), which is a common situation. As this means it is not possible to simply increase time in one behaviour - it is only possible to spend time in one behaviour rather than in another behaviour - it is the relative values that are relevant. 
```{r}
head(apply(simdata[, c("vigorous", "moderate", "light", "sedentary", "sleep")], 1, sum))
```
Other examples of compositional data don't have a fixed sum. For example, consider the vote counts in an election for different parties in different constituencies: although the overall number of votes cast might differ between constituencies, it is the proportion of the vote going to each party, and not the total number of votes going to that party in a given constituency, that is relevant. 

#### Notes
* The variables in compositional data are often known as *parts*.
* In Compositional Data Analysis, expressing the data as proportions of a whole is known as taking the *closure*, written and defined as: 
\begin{equation}
  \mathcal{C}(part_{1}, part_{2}, ..., part_{D}) = \Big(\frac{part_{1}}{\sum_{i =1}^{D}{part_i}}, \frac{part_{2}}{\sum_{i =1}^{D}{part_i}}, ..., \frac{part_{D}}{\sum_{i =1}^{D}{part_i}}\Big)
\end{equation}

## Exploratory data analysis

### The compositional mean
The compositional mean is a frequently-used summary statistic for compositional data [@Filzmoser2018]. Like the standard (multivariate) mean it is a measure of location, or central tendency. Unlike the standard mean, it interacts in an appropriate way with compositional data. 

To calculate the compositional mean of columns `"vigorous", "moderate", "light", "sedentary", "sleep"` using `epicoda`, we run: 
```{r}
comp_mean(data = simdata, # this is the dataset
          comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep") # this is the labels of the compositional columns, 
          # which we specified above
          )
```
There are a few things to notice about the output. Even though we started out with sums which added to 24, the mean seems to add to 1. This is because we didn't specify the units - and so the package didn't know what units to return it to us in. But it could still return it as a sum to 1, without loss of information, as per the definition of compositional data. 

We've also got a message about zeroes. Compositional data analysis can't automatically include zero values. For more information about this issue, see [@Palarea-Albaladejo2015] and [@Rasmussen2020]. There are a couple of options: 

* They can be dropped (this is the right option if they're so called 'true zeroes', and not related to a limit in the precision with which we can measure). To drop them, set `rounded_zeroes = FALSE`. 

* They can be imputed. This is the right option if they're so-called 'rounded zeroes' i.e. they're related to a limit in the precision with which we can measure. For example, with activity data, it might be that if we recorded data for long enough, we would eventually see some vigorous activity, but our measurement period wasn't long enough. It might also be that our measurement device can't record values lower than a certain level.

To impute the zeroes, as per the second option, we need to specify the minimum value detectable in the data. This can be specified using the `det_limit` argument, and it should be specified with the same scale as the compositional columns in the input data. [Note that if the input data scale and the output scale (`units` argument), are different, it should match *the input data scale*.]

The default in the function, if neither `rounded_zeroes` or `det_limit` is specified, is to impute them with the detection limit (`det_limit`) set as the minimum non-zero value observed in the data. This isn't necessarily a correct default!

So we can now find the mean again, this time imputing the zeroes with a specified detection limit and specifying the units.  In the simulated data, the shortest period of a behaviour we could detect was $0.0012\ hr/day$ (due to the length of an epoch for the accelerometer used to measure the data on which the simulated data is based). 

```{r}
comp_mean(data = simdata, # this is the dataset
          comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"), # this is the labels of the compositional columns, 
          # which we specified above
          rounded_zeroes = TRUE, # this option specifies that we'll treat the zeroes 
          # as rounded zeroes i.e. we'll impute them
          det_limit = 0.0012, # this is the smallest value observable in the data
          units = "hr/day" # this is the units. There are pre-specified options "hr/day",
          # "hr/wk", "min/day", "min/wk" and "unitless". 
          # If you set units = "specified", you can also specify your own units using 
          # specified = c("my_units_name", sum of a composition in these units)
          )
```

Many other functions in the package have `rounded_zeroes` and `det_limit` arguments, because they use calculations which need zeroes to be imputed or excluded. For all of them, it works the same as for `comp_mean`. For the simulated data, the default is fine, so in the remainder of this vignette they are left at the default. 

### Other descriptive statistics
While the compositional mean is a fairly clear compositional equivalent to the mean for non-compositional data, other standard descriptive statistics (such as the variance), do not have straightforward compositional equivalents for all purposes for which they are used (there may be several alternatives, depending on the purpose of the analysis) [@Filzmoser2018]. These are beyond the current scope of this package. 

### Plotting ternary diagrams
Ternary plots provide a way to plot compositional data with three parts. The R package `ggtern` is an extension of `ggplot2` to carry out this plotting. In this package, there is a wrapper for `ggtern` functions to implement common use cases in epidemiological analyses: plotting the data density, and plotting confidence regions for the data, both overall and within subgroups.

**Note (18.03.2021):** Currently, a conflict between `ggtern` and `ggplot2` means using `ggtern` can stop axis labels appearing on `ggplot2` graphs. Pending a more permanent solution, these functions remain in the package, but `ggtern` is not automatically loaded and should be used with care. They are not run in this vignette.



Frequently, data has more than three compositional parts (as here). There are two alternatives in that case: using three parts of the existing parts (looking at the subcomposition formed by these parts), or summing parts to consider a different composition (where that is informative). In the simulated data, `"vigorous", "moderate"` and `"light"` represent three different intensities of activity, so it is meaningful to combine them: 
```{r, eval = FALSE}
simdata$all_activity <- simdata$vigorous + simdata$moderate + simdata$light
plot_confidence_region_ternary(data = simdata, 
                     parts_to_plot = c("all_activity", "sedentary", "sleep"),
                     probs = c(0.25, 0.5, 0.75) # This will plot 25, 50 and 75% confidence regions (default is 0.5, 0.9, 0.95)
                      )

```

Another frequent use case is considering this plot separated by subgroups. For example, to look at different sexes (which in fact don't differ much in the simulated data):  
```{r, eval=FALSE}
plot_confidence_region_ternary(data = simdata, 
                     parts_to_plot = c("all_activity", "sedentary", "sleep"),
                     groups = "sex", 
                     probs = c(0.25, 0.5, 0.75) # This will plot 25, 50 and 75% confidence regions (default is 0.5, 0.9, 0.95)
                      )

```

The `mark_points` argument can be used to annotate particular points using crosshairs. For example, to mark the compositional mean: 
```{r,eval = FALSE}
points_to_mark <- as.data.frame(comp_mean(data = simdata, comp_labels = c("all_activity", "sedentary", "sleep")))
plot_confidence_region_ternary(data = simdata, 
                     parts_to_plot = c("all_activity", "sedentary", "sleep"),
                     mark_points = points_to_mark,
                     probs = c(0.25, 0.5, 0.75) # This will plot 25, 50 and 75% confidence regions (default is 0.5, 0.9, 0.95)
                      )                              
```


## Regression modelling 

### Setting up regression models 
The package has a wrapper for linear regression, logistic regression and Cox regression modelling using compositional variables, with adjustment for covariates. Compositional Data Analysis regression modelling works by transforming the variables making up the composition (see 'Mechanics of Compositional Data Analysis' below), and using the transformed variables (known as 'log-ratio coordinates') in the model. The package automates these steps. For example, to set up a linear regression model for the association between behaviour variables and `BMI`, adjusted for age and sex: 
```{r}
lm_BMI <- comp_model(type = "linear",
         outcome = "BMI",
         covariates = c("agegroup", "sex"),
         data = simdata,
         comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
```

Logistic models can be produced in exactly the same way by setting `type = "logistic"` and using an appropriate binary outcome variable. Cox models can be produced similarly by setting `type = "cox"` and either setting `outcome` as a `Surv` object from the package `survival` or setting the `follow_up_time` and `event` arguments. For examples, see 'Regression modelling: further examples'.

### Understanding the model output
The model object is a standard `lm` object, and we can inspect it in the usual way:
```{r}
plot(lm_BMI)
summary(lm_BMI)

```

The model diagnostics can be interpreted as usual, and in this case all looks fine. However, the summary is hard to interpret because the transformed variables or 'coordinates' used in the model do not relate in a straightforward way to the original composition. 

### Presenting the results: tabulate pivot-coordinate coefficients
A common way to present the results is to present coefficients as usual for covariates, and then produce as many models as there are compositional variables. From each of these models, the coefficient for the first 'ilr pivot coordinate' is taken (see 'Mechanics of Compositional Data Analysis' below). These ilr pivot coordinate coefficients can be interpreted in terms of reallocation between the particular compositional part and all other parts proportionally. `tab_coefs` is a function which wraps generating one model per compositional variable and outputting a table of model coefficients. 
```{r}
tab_coefs(scale_type = "lp", # This argument can be "lp" or "exp" and determines whether 
          # coefficients are presented on the scale of the linear predictors ("lp") 
          # or are exponentiated ("exp"). Exponentiation gives the Odds Ratio for 
          # logistic regression models and the Hazard Ratio for Cox regression models. 
          level = 0.95, 
          type = "linear",
          outcome = "BMI",
          covariates = c("agegroup", "sex"),
          follow_up_time = "follow_up_time",
          event = "event",
          data = simdata,
          comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))

```
#### Notes: 
* As these coefficients come from different models they do not parametrise a single model. 
* Arguments which would be needed for `comp_model` must be given to this function too, as it regenerates the regression models.
* The function `tab_covariate_coefs` is called within this function, and can be used to extract the coefficients of the covariates only from a model. 

### Presenting the results:  substitution plots
Substitution or reallocation plots show the model-estimated difference in the outcome (relative to at the compositional mean) as the balance between two parts of the composition varies. In the case of time use/physical behaviours, they are sometimes called isotemporal substitution plots. They can be thought of as showing the difference in outcome associated with a hypothetical reallocation of time from one behaviour to another, as estimated by the model. For example: 
```{r}
plot_transfers(from_part = "sedentary",
               to_part = "moderate",
               model = lm_BMI ,
               comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"),
               y_label = "Model-estimated difference in BMI",
               units = "hr/day")
```

#### Notes 
* To customise the appearance of the plot see 'Customising plots'. 
* Results of analyses using a Compositional Data Analysis approach can be presented using reallocations other than 1-1 between two parts. This is not currently implemented in the `plot_transfers` function. However, the `predict_fit_and_ci` function can be used to estimate the outcome for any specified reallocation. 
* There is more information on how confidence intervals for this plot are calculated in a vignette (`vignette("derivation_of_CIs_used")`).

### Presenting the results: estimates at particular compositions
Another set of estimates that might be of interest are estimates at particular compositions. These can be examined by plotting these in a 'forest plot' style plot. This allows assessment of the model-estimated associations with much more complex reallocations between parts. 

#### Generating perturbations of a composition
While there are hugely varied compositions which might be possible, one kind of composition is often of interest. This is compositions where one particular part is perturbed at the expense of all other parts proportionally. As this is a common use case, there is a function implemented to do this: 
```{r} 
df <- as.data.frame(
  comp_mean(
    data = simdata,
    comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"), 
    units = "hr/day" # Note whatever the units are here is the units 
    # you should specify the changes in in the next two lines. 
  )
)
new_comp <-
  change_composition(
    composition = df,
    main_part = "moderate",
    main_change = +0.5,
    comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep")
  ) 

new_comp2 <-
    change_composition(
    composition = df,
    main_part = "sedentary",
    main_change = -3.5,
    comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep")
  )
list_for_plot <- list("Extra 0.5 hr/day moderate" = new_comp, "3.5 hr/day less sedentary" = new_comp2)
print(list_for_plot)
```

#### Plotting estimates at particular compositions
We now need to plot these, which can be done as follows. 
```{r}
forest_plot_comp(composition_list = list_for_plot, 
                 model = lm_BMI, 
                 comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
```

The `forest_plot_comp` function can also be used to compare models. To do this, leave `model` at its default `NULL` value, and set `models_list` as a named list of models. For example, to compare `lm_BMI` with a model adjusted for `age_group` only and an unadjusted model (very boring in this case as the simulated data doesn't have realistic correlation structure for covariates!): 
```{r}
lm_BMI_unadjusted <- comp_model(type = "linear",
         outcome = "BMI",
         data = simdata,
         comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
lm_BMI_age_group_only <- lm_BMI <- comp_model(type = "linear",
         outcome = "BMI",
         covariates = c("agegroup"),
         data = simdata,
         comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
forest_plot_comp(composition_list = list_for_plot, 
                 models_list = list("Unadjusted" = lm_BMI_unadjusted, 
                                    "Age-adjusted" = lm_BMI_age_group_only,
                                    "Age- and sex- adjusted" = lm_BMI), 
                 comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
```

## Additional information 
This section contains lots of additional information, on: 

* Mechanics of Compositional Data Analysis: Log-ratio transformations
* Presenting results via model predictions 
* Customising plots 
* Regression modelling: further examples
* Getting the numbers out: estimates/ predictions for particular compositions

### Mechanics of Compositional Data Analysis: Log-ratio transformations
The functions above wrap all the 'log-ratio mechanics' of using a Compositional Data Analysis approach.

There are several different kinds of transformations which can be used in order to work with compositional data (alr, ilr, clr). These transform the variables into coordinates which are capable of representing all the variation in the data. Different transformations have different pros/cons, and alr and clr can only be used for certain applications. All the functions default to ilr (isometric log ratio) transformation using so-called 'pivot coordinates', which is fairly standard in the physical behaviour/time use literature (and is appropriate for the applications which the other transformations are used for). If you want to learn more, see [@Filzmoser2018][@Pawlowsky-Glahn2011].

The purpose of this package is not to provide this functionality, which is implemented in other R packages, such as `compositions` and `robCompositions`. (The main aim of this package is shown above, in the functions to perform standard epidemiological analyses). This functionality is reimplemented here simply to ensure that future changes to the defaults in those packages do not affect this package.
```{r}
data_ilr_impute_zeroes <- transform_comp(data = simdata,
                                   comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"),
                                   transformation_type = "ilr")
print(head(data_ilr_impute_zeroes))
```
There are also a few other things we can do with this function, like specifying a given 'first part' (which is useful when it comes to reporting, and is used therefore in `tab_coefs`). To learn more, run `?transform_comp`. 

### Presenting results via model predictions 
An alternative which may sometimes be useful when interpreting models is to consider model-predicted outcomes for different input variable values, including values of covariates (rather than model-estimated differences associated with a particular difference in compositional variables). Functions related to plotting (e.g. the `plot_transfers` function) can do this by setting `terms = FALSE`: 
```{r}
plot_transfers(from_part = "sedentary",
               to_part = "moderate",
               model = lm_BMI ,
               comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"),
               y_label = "Model-predicted outcome",
               units = "hr/day",
               terms = FALSE)
```

#### Notes
* The predictions and confidence intervals on this plot capture the predictions and uncertainty on them due to all variables in the model, including the covariates at the set fixed values. 
* The `fixed_values` argument can be used to set the values at which covariates are fixed (which if not set default to median or modal values for continuous and categorical variables respectively)
* Other functions, such as `forest_plot_comp`, can also output predictions by setting `terms = FALSE`. 

### Customising plots 
All the plotting functions can have settings changed to make more visually attractive plots (e.g. changing text size, line size, text colour). `plot_transfers` and `simplex_plotting` are based on `ggplot2` and its extension `ggtern`. 

* To specify non-data aspects of the plot (e.g. line, axis and text size and colours), set the optional `theme` argument. Note that you will need the `ggplot2` package loaded in order to do this.
* For `plot_transfers`, to specify characteristics of the plotted line, set the `point_specification` argument as a `geom_point` object.
* For `plot_transfers`, to specify colours of the error bars, set the `error_bar_colour` argument as the name of a colour.
* For `plot_transfers`, if the line looks 'gappy' (likely at the ends), increase `granularity` (and if it takes too long to plot, decrease `granularity`). 

For an extreme example of implementing some of this: 
```{r}
library(ggplot2) # it's easier to specify e.g. themes if ggplot2 is loaded. 
plot_transfers(from_part = "sedentary",
               to_part = "moderate",
               model = lm_BMI ,
               comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"),
               y_label = "Model-estimated difference in outcome",
               units = "hr/day",
               point_specification = geom_point(size = 0.5, colour = "purple"),
               error_bar_colour = "yellow", 
               theme = theme_bw()+ theme(line = element_line(colour = "red", size = 2), 
                             text = element_text(family = "mono", face = "bold.italic", colour = "green"), 
               axis.line = element_line(colour = "blue", size = 3)))
```

Slightly differently, `forest_plot_comp` is based on the `forestplot` package, so takes arguments specifying the plot in the same way as that function. There are several arguments which set particular aspects of the forest plot. In addition, any argument that the basic `forestplot` function can take, which isn't already set, can be set as an argument to `forest_plot_comp`.
```{r}
library(forestplot) # it's easier to specify options such as fpColors if the forestplot package is loaded. 
forest_plot_comp(composition_list = list_for_plot, 
                 model = lm_BMI, 
                 comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"), 
                 pred_name = "The estimated difference at this composition", # This sets the name with which to label the column. , 
                 title = "Analysis of physical behaviour data", # This sets the overall plot title. 
                 col = forestplot::fpColors(zero = "black") # This sets the colour of the line at 0 (or 1, if it's the plot of a logistic or Cox model)
                 )

```

For additional information on how to customise the functions, see the manual by running e.g. `?plot_transfers`. 

### Regression modelling: further examples

As well as linear models, we can produce logistic and Cox models:
```{r}
log_outcome <- comp_model(type = "logistic",
         outcome = "disease",
         covariates = c("agegroup", "sex"),
         data = simdata,
         comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
cox_outcome <- comp_model(type = "cox",
         event = "event", 
         follow_up_time = "follow_up_time",
         covariates = c("agegroup", "sex"),
         data = simdata,
         comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
```


### Getting the numbers out: predictions for particular compositions
To get the numbers out for predictions at particular compositions, we can use the `predict_fit_and_ci` function (which is wrapped in the plotting functions).This takes a data frame argument for `new_data`. 

```{r}
predict_fit_and_ci(model = lm_BMI, new_data = new_comp, comp_labels = c("vigorous", "moderate", "light", "sedentary", "sleep"))
```

As noted above, this can be used to look at any reallocations between parts which are of interest. 

# References
